---
title: 'Second Analysis: Replication Desmond 2015'
author: "Jan Pfänder"
date: "2023-05-23"
output:
  html_document:
    keep_md: yes
  pdf_document: default
---

```{r packages, message=FALSE, warning=FALSE}
# load required packages
library("tidyverse")
library("readxl") 
library("mice")
```

The instructions for the second analysis are:

> *"Your analysis should produce a single, main result in terms of statistical families of z-, t-, F-, or χ² tests (or their alternative or non-parametric versions). \n You should control for Black Renter, Hispanic Renter, Other Ethnicity Renter Age, Less Than High School Education,High School/GED, Housing Assistance in Past Residence Single-Mother Household in Past Residence Criminal Record Before Move ,Had a Child in Previous 2 Years"*

Following these instructions, I will proceed as for the first analysis, but this time *leaving out* the following control variables:

- relationship_dissolution 
- job_loss 
- benefits_loss 
- poverty_previous

### Read cleaned data

```{r, message=FALSE}
# export cleaned file
data <- read_csv("data/cleaned_data_combined.csv") %>% 
  mutate(move_reason = relevel(as.factor(move_reason), ref = "voluntary/NA"))
```

### Impute NAs

```{r}
#impute NAs
# make various samples
data_imputed_NAs <- mice(data,
                m=15,
                seed=12345)
```

### Analyse

Run final model with imputed values
```{r}
models <- with(data_imputed_NAs, 
               lm(poverty_current ~ move_reason +
                    race +
                    age +
                    education_less_than_highschool +
                    education_highschool +
                    housing_assistance +
                    single_mom +
                    criminal_record +
                    recent_child, 
                  weights=weight))

result <- pool(models) %>% 
  tidy() %>% 
  mutate_if(is.numeric, round, digits = 4)
```

### Results

#### Table

```{r}
result
```

#### Main result

```{r}
result %>% filter(term == "move_reasonforced") %>% 
  select(term, estimate, std.error, statistic, p.value)
```




